Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-04-07 paper.tex
figures,bib,txt

{\it Spatial distribution -- stars and dust}\\ Our goal in this
section is to derive a simple spatial model for the distribution of SD
in galaxy clusters. For this reason we use stacked cluster
observations by Sloan in the $r$-band that trace the stars. We assume
that the dust trace the stars in the clusters, which should be
accurate to the order of magnitude estimate that we are interested in
here. The stacked surface brightness profiles in
\cite{2005MNRAS.358..949Z} are measured at redshift $z \sim 0.25$ with
an average mass of the clusters of $4.0\times10^{13}\,\msun$. The
profiles are build up out of three components; diffuse intracluster
light (ICL), galaxies, and the brightest cluster galaxy (BCG) in the
center of clusters. However, the total volume overlap of the ISM to
the cluster is very small compared to the ICL- and BCG overlap with
the cluster, thus the relative contribution of the galaxies to the IC
emission is suppressed. To remove this bias we cut out the galaxies
that have been smoothed over the cluster in the stacked analysis. In
Fig.~\ref{fig:SD_spatial} we show the SDSS stacked brightness profiles
as well as the fitted profiles\footnote{The measured brightness is
  converted into units of $\rmn{erg}\,\rmn{s}^{-1}$ using
  \cite{2010...book} $S(\rmn{mag}\,''^{-2}) =
  \mathcal{M}_\odot+21.572-2.5\rmn{log}_{10}S(L_\odot\,\rmn{pc}^{-2})$,
  where the sun's absolute magnitude in the $r$-band is given by
  $\mathcal{M}_\odot=27.1$ \cite{2000asqu.book..339L} and the
  luminosity of the sun by $L_\odot=3.85\,10^{33}
  \rmn{erg}\,\rmn{s}^{-1}$.}, both scaled with the unitless $\rvir$
normalized area to resembele the expected luminosity at radius
$r$. Our benchmark spatial model is shown with the solid black line,
and include the contribution from the ICL and BCG.

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/SB.photon.eps}
\caption{\it Spatial dependence of light from stars and dust. We show
  2D surface brightness profiles in the $r$-band ($\sim 1\,\ev$)
  obtained from stacked clusters in the Sloan Digital Sky Survey
  (SDSS) at the redshift $z \sim 0.25$ \cite{2005MNRAS.358..949Z}. The
  brightness profiles have been weighted with the $\rvir$ normalized
  area inside radius $r$, and trace the radial dependence of the
  luminosity from stars and galaxies. The crosses show the total
  observed starlight including the diffuse intracluster (ICL), the
  galaxies, and the brightest cluster galaxy (BCG) in the center of
  the cluster. The diamonds denote the observed starlight from the
  ICL and the BCG. The solid line show the fit to the data of the
  total light including the ICL, the BCG, and the galaxies, while the
  smooth component is represented by the dashed line that is fitted to
  only the ICL and the BCG.}
 \label{fig:SD_spatial}
\end{figure}

Instead of modelling the surface brightness with a de Vaucouleur
profile with the addition of a power-law, we use a double beta profile
model for simplicity of deprojection given by:
\begin{equation}
S_\sd (r_\bot)= \sum_{i=1}^2 \,S_i\, 
\left[1 + \left( \frac{r_\bot}{r_{\mathrm{c}_i}}\right)^2\right]
^{-3\beta_i + 1/2}\,,
\label{double_beta}
\end{equation}
where we found the following best fitted central brightness ($S_i$),
core radius ($r_{\mathrm{c}_i}$), and slope ($\beta_i$):
\begin{eqnarray}
 S_1&=&1.4\times10^{-5}\,\rmn{erg}\,\rmn{cm}^{-2}\,\rmn{s}^{-1},\,
r_{\mathrm{c}_1}=210\,\rmn{kpc},\,
\beta_1=0.44\nonumber\\
 S_2&=&6.0\times10^{-3}\,\rmn{erg}\,\rmn{cm}^{-2}\,\rmn{s}^{-1},\,
r_{\mathrm{c}_2}=2.8\,\rmn{kpc},\,
\beta_2=0.45\,.\nonumber\\
\label{fit_spatial_IR}
\end{eqnarray}
The 3D spatial profile is derived by deprojecting the surface
brightness in Eq.~(\ref{double_beta}) (see
e.g. \cite{2004A&A...413...17P} for details about the deprojection):
\begin{eqnarray}
  j(r)  = \sum_{i=1}^2 \frac{S_i}{2\pi\,r_{\mathrm{c}_i}}\,
  \frac{6 \beta_i - 1}{\left(1 + r^2/r^2_{\mathrm{c}_i}\right)^{3\beta_i}}\,
  \mathcal{B}\left(\frac{1}{2},3\beta_i\right)\,,
\end{eqnarray}
where $\mathcal{B}(a,b)$ denotes the beta-function
\cite{1965hmfw.book.....A}.

In Fig.~\ref{fig:SD_spectra} we show the spectral distribution of the
light from SD together with the CMB black-body distribution in a
galaxy cluster, both at the radius $r=0.071\rvir$ where the energy
density of the smooth light from SD (see Fig.~\ref{fig:SD_Edens})
equals the energy density of the CMB. We have renormalized both the
spectral data and the fitted spectra using the luminosity from stars
and dust individually (see the normalization section below for further
details).
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/fit.porter.v2.eps}
\caption{\it Spectral dependence of radiation fields in a cluster of
  galaxies. The black dotted line in the left peak show the spectrum
  of CMB photons using a black body with a temperature of
  $2.73$~K. The crosses in the middle and right peaks represent the
  measured spectra from stars and dust (SD), respectively, and are
  derived in \cite{2006ApJ...648L..29P} for a galaxy. We normalize the
  individual SD spectrum separately using the observed luminosity from
  SD in clusters. The SD luminosity is related to the cluster mass
  through Eq, where we use have used the mass
  $\mvir=4.0\times10^{13}\msun$ in this figure. Finally we renormalize
  the SD spectra to the radius $r=0.071\rvir$, where the smooth energy
  density of the SD light (see Fig.~\ref{fig:SD_Edens}) equals the
  energy density of the CMB. The red dashed lines show the fitted SD
  spectral model.}
 \label{fig:SD_spectra}
\end{figure}
We choose to fit the spectrum of the dust that peaks at about
$10^{-2}\,\ev$ with a double power-law, and the slightly wider
spectral distribution of the stars that peaks at about $1\,\ev$ with a
tipple power-law. We find the best fit spectra for the SD in a
cluster through:
\begin{eqnarray}
  \eph^2\frac{\dd^2 N_\ph}{\dd \eph \dd V}
  &\equiv& u_\sd(\eph) =  \sum_i N_i(\lx) u_i^\gal(\eph)\,,\nonumber \\ 
\rmn{where}\quad i&=&\{\rmn{stars,dust}\} \qquad \rmn{and}\\ 
% &=&  N_\dust(\lx) u_\dust^\gal(\eph) + N_\stars(\lx) u_\stars^\gal(\eph)  \nonumber \\ 
%  \\
  u_\stars^\gal(\eph) &=& \frac{23\,\rmn{eV}}{\rmn{cm}^3} 
  \left(\frac{1.23\,\rmn{eV}}{\eph}\right)^{1.9} \nonumber \\
  &\times&\left[1+\left(\frac{2.04\,\rmn{eV}}{\eph}\right)^{20}\right]
  ^{-\frac{1.9}{20}}\nonumber \\
  &\times& \left[1+\left(\frac{0.78\,\rmn{eV}}{\eph}\right)^{20}\right]^{-\frac{2.6}{20}}\,, \\
  u_\dust^\gal(\eph) &=& 
  \frac{40\,\rmn{eV}}{\rmn{cm}^3} 
  \left(\frac{0.0144\,\rmn{eV}}{\eph}\right)^{4.9}\nonumber \\
  &\times& \left[1+\left(\frac{0.0144\,\rmn{eV}}{\eph}\right)^{1.9}\right]^{-4.9}\,.
\end{eqnarray}

The normalization constant for the starlight $N_\stars(\lx)$ is derived
in the following three steps: (1) We use the average apparent magnitude
in the $(r+i)$-band $m_{(r+i),0.25}\approx 15.5$
\cite{2005MNRAS.358..949Z} of a cluster with the average mass
$\mvir=4.0\times10^{13}\msun$ at a redshift $z=0.25$ to derive the
luminosity from the stars:
\begin{equation}
L_\stars=10^{\frac{m_\odot-m_{(r+i),0.25}}{2.5}}
\left(\frac{D_\clu}{D_\odot}\right)^2 L_\odot
\approx 7.1\times10^{44}\,\rmn{erg}\,\rmn{s}^{-1}\,,
\label{eq:L_star}
\end{equation}
where the apparent magnitude of the sun $m_\odot\approx -27.1$, the
distance to the sun $D_\odot=4.85\times10^{-6}\,\rmn{pc}$, the
luminosity of the sun $L_\odot=3.85\times10^{34}\,\rmn{erg\,s}^{-1}$,
and the distance to the cluster $D_\clu=1.26\times10^9\,\rmn{pc}$. (2)
We use the IR to X-ray scaling relation in Eq.~(\ref{eq:SD-X}) to derive
a more general form of the luminosity from stars in
Eq.~(\ref{eq:L_star}) that is valid for an arbitrary cluster mass. The
cluster X-ray luminosities are found in the extended HIFLUGCS
catalogue \cite{2002ApJ...567..716R} to which we have added the Virgo
cluster. (3) The normalization constant for the stars is
determined from the above steps together with Eq.~(\ref{eq:E_SD})
where we integrate over the cluster volume and spectral distribution
of the stars. It is given by
\begin{equation}
 N_\stars(\lx) = 1.2\times10^{-9}\,
\left(\frac{\lx}{10^{44}\rmn{erg\,s}^{-1}}\right)^{0.53}\,.
\label{eq:N_stars}
\end{equation}
Note that since the luminosities include the contribution from
galaxies, we have integrated the spatial distribution including
galaxies to derive the normalization. Once the normalization is fixed,
the SD energy densities are derived from the spatial distribution
where galaxies have been excluded.

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/ucool.eps}
 \caption{\it Spatial dependence of the energy density of radiation
   fields in a cluster of galaxies. The energy density of the CMB,
   shown by the blue dotted line, is isotropic throughout the cluster,
   hence represented by a flat profile with
   $u_\rmn{cmb}=0.26\,\rmn{eV}\,\rmn{cm}^{-3}$. The energy density of
   the light from stars and dust (SD) is denoted by the red dashed
   line and the solid purple line for the total SD light and the
   smooth SD light, respectively. The SD light has been renormalized
   to a cluster with the mass $\mvir=4.0\times10^{13}\msun$. Finally we
   show the energy density two magnetic field models with a central
   magnetic field of 3~$\mu G$. The magnetic field scales with the gas
   density to the power $\alpha_\rmn{B}$; green dash-dotted line
   ($\alpha_\rmn{B}=0.5$) and orange dash-dotted line
   ($\alpha_\rmn{B}=0.7$). Note that the SD radiation is dominating
   the energy density inside $\sim0.1\,\rvir$.}
 \label{fig:SD_Edens}
\end{figure}

\del{In this section we present the formalism for the different
galaxy cluster emission components and show their contribution to the
total dark matter gamma-ray spectrum. We account for the emission
radiated both by final state radiation and inverse Compton where we
allow for upscattering through CMB, starlight, and dust.}

%Using a typical
%X-ray luminosity of $L_\rmn{X}=10^{44}\,\rmn{erg\,s}^{-1}$ results in a
%IR energy density
%$N\,E_\IR\,f_\rmn{tot}(E_\IR)\,j(20\,\rmn{kpc})=u_\IR\simeq 4\,\rmn{erg\,
%  cm}^{-1}$ in the region around 20~kpc from the cluster center at the
%energy of the maximum IR energy density $E_\IR=0.011\,\rmn{eV}$.

@@

\begin{figure}
  \includegraphics[width=0.99\columnwidth]{figures/MLscaling.eps}
  \caption{\it Lum-scaling relations, mass dependence of SD cooling
    have been neglected, actually a factor few difference for
    groups. factor 2 for Fornax, and small difference for large
    clusters}
 \label{fig:lum_mass_scaling}
\end{figure}

@@

>>>>

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/SB.photon.eps}
\caption{\it Spatial dependence of light from stars and dust. We show
  2D surface brightness profiles in the $r$-band ($\sim 1\,\ev$)
  obtained from stacked clusters in the Sloan Digital Sky Survey
  (SDSS) at the redshift $z \sim 0.25$ \cite{2005MNRAS.358..949Z}. The
  brightness profiles have been weighted with the $\rvir$ normalized
  area inside radius $r$, and trace the radial dependence of the
  luminosity from stars and galaxies. The crosses show the total
  observed starlight including the diffuse intracluster (ICL), the
  galaxies, and the brightest cluster galaxy (BCG) in the center of
  the cluster. The diamonds denote the observed starlight from the
  ICL and the BCG. The solid line show the fit to the data of the
  total light including the ICL, the BCG, and the galaxies, while the
  smooth component is represented by the dashed line that is fitted to
  only the ICL and the BCG.}
 \label{fig:SD_spatial}
\end{figure}
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/fit.porter.v2.eps}
\caption{\it Spectral dependence of radiation fields in a cluster of
  galaxies. The black dotted line in the left peak show the spectrum
  of CMB photons using a black body with a temperature of
  $2.73$~K. The crosses in the middle and right peaks represent the
  measured spectra from stars and dust (SD), respectively, and are
  derived in \cite{2006ApJ...648L..29P} for a galaxy. We normalize the
  individual SD spectrum separately using the observed luminosity from
  SD in clusters. The SD luminosity is related to the cluster mass
  through Eq, where we use have used the mass
  $\mvir=4.0\times10^{13}\msun$ in this figure. Finally we renormalize
  the SD spectra to the radius $r=0.071\rvir$, where the smooth energy
  density of the SD light (see Fig.~\ref{fig:SD_Edens}) equals the
  energy density of the CMB. The red dashed lines show the fitted SD
  spectral model.}
 \label{fig:SD_spectra}
\end{figure}
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/ucool.eps}
 \caption{\it Spatial dependence of the energy density of radiation
   fields in a cluster of galaxies. The energy density of the CMB,
   shown by the blue dotted line, is isotropic throughout the cluster,
   hence represented by a flat profile with
   $u_\rmn{cmb}=0.26\,\rmn{eV}\,\rmn{cm}^{-3}$. The energy density of
   the light from stars and dust (SD) is denoted by the red dashed
   line and the solid purple line for the total SD light and the
   smooth SD light, respectively. The SD light has been renormalized
   to a cluster with the mass $\mvir=4.0\times10^{13}\msun$. Finally we
   show the energy density two magnetic field models with a central
   magnetic field of 3~$\mu G$. The magnetic field scales with the gas
   density to the power $\alpha_\rmn{B}$; green dash-dotted line
   ($\alpha_\rmn{B}=0.5$) and orange dash-dotted line
   ($\alpha_\rmn{B}=0.7$). Note that the SD radiation is dominating
   the energy density inside $\sim0.1\,\rvir$.}
 \label{fig:SD_Edens}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=0.99\columnwidth]{figures/MLscaling.eps}
  \caption{\it Luminosity to mass scaling relations. We show the
    luminosity from; CRp induced emission (red solid), a leptophilic
    model that IC upscatter CMB photons (green dash-dotted), mass
    dependence of SD cooling have been neglected, actually a factor
    few difference for groups. factor 2 for Fornax, and small
    difference for large clusters}
 \label{fig:lum_mass_scaling}
\end{figure}

@@

{\it Spatial distribution -- stars and dust}\\ Our goal in this
section is to derive a simple spatial model for the distribution of SD
in galaxy clusters. For this reason we use stacked cluster
observations by Sloan in the $r$-band that trace the stars. We assume
that the dust trace the stars in the clusters, which should be
accurate to the order of magnitude estimate that we are interested in
here. The stacked surface brightness profiles in
\cite{2005MNRAS.358..949Z} are measured at redshift $z \sim 0.25$ with
an average mass of the clusters of $4.0\times10^{13}\,\msun$. The
profiles are build up out of three components; diffuse intracluster
light (ICL), galaxies, and the brightest cluster galaxy (BCG) in the
center of clusters. However, the total volume overlap of the ISM to
the cluster is very small compared to the ICL- and BCG overlap with
the cluster, thus the relative contribution of the galaxies to the IC
emission is suppressed. To remove this bias we cut out the galaxies
that have been smoothed over the cluster in the stacked analysis. In
Fig.~\ref{fig:SD_spatial} we show the SDSS stacked brightness profiles
as well as the fitted profiles\footnote{The measured brightness is
  converted into units of $\rmn{erg}\,\rmn{s}^{-1}$ using
  \cite{2010...book} $S(\rmn{mag}\,''^{-2}) =
  \mathcal{M}_\odot+21.572-2.5\rmn{log}_{10}S(L_\odot\,\rmn{pc}^{-2})$,
  where the sun's absolute magnitude in the $r$-band is given by
  $\mathcal{M}_\odot=27.1$ \cite{1998gaas.book.....B} and the
  luminosity of the sun by $L_\odot=3.85\,10^{33}
  \rmn{erg}\,\rmn{s}^{-1}$.}, both scaled with the unitless $\rvir$
normalized area to resembele the expected luminosity at radius
$r$. Our benchmark spatial model is shown with the solid black line,
and include the contribution from the ICL and BCG.

Instead of modelling the surface brightness with a de Vaucouleur
profile with the addition of a power-law, we use a double beta profile
model for simplicity of deprojection given by:
\begin{equation}
S_\sd (r_\bot)= \sum_{i=1}^2 \,S_i\, 
\left[1 + \left( \frac{r_\bot}{r_{\mathrm{c}_i}}\right)^2\right]
^{-3\beta_i + 1/2}\,,
\label{double_beta}
\end{equation}
where we found the following best fitted central brightness ($S_i$),
core radius ($r_{\mathrm{c}_i}$), and slope ($\beta_i$):
\begin{eqnarray}
 S_1&=&5.0\times10^{-6}\,\rmn{erg}\,\rmn{cm}^{-2}\,\rmn{s}^{-1},\,
r_{\mathrm{c}_1}=130\,\rmn{kpc},\,
\beta_1=0.54\nonumber\\
 S_2&=&3.9\times10^{-3}\,\rmn{erg}\,\rmn{cm}^{-2}\,\rmn{s}^{-1},\,
r_{\mathrm{c}_2}=2.8\,\rmn{kpc},\,
\beta_2=0.53\,.\nonumber\\
\label{fit_spatial_IR}
\end{eqnarray}
The 3D spatial profile is derived by deprojecting the surface
brightness in Eq.~(\ref{double_beta}) (see
e.g. \cite{2004A&A...413...17P} for details about the deprojection):
\begin{eqnarray}
  j(r)  = \sum_{i=1}^2 \frac{S_i}{2\pi\,r_{\mathrm{c}_i}}\,
  \frac{6 \beta_i - 1}{\left(1 + r^2/r^2_{\mathrm{c}_i}\right)^{3\beta_i}}\,
  \mathcal{B}\left(\frac{1}{2},3\beta_i\right)\,,
\end{eqnarray}
where $\mathcal{B}(a,b)$ denotes the beta-function
\cite{1965hmfw.book.....A}.

In Fig.~\ref{fig:SD_spectra} we show the spectral distribution of the
light from SD together with the CMB black-body distribution in a
galaxy cluster, both at the radius $r=0.071\rvir$ where the energy
density of the smooth light from SD (see Fig.~\ref{fig:SD_Edens})
equals the energy density of the CMB. We have renormalized both the
spectral data and the fitted spectra using the luminosity from stars
and dust individually (see the normalization section below for further
details).

We choose to fit the spectrum of the dust that peaks at about
$10^{-2}\,\ev$ with a double power-law, and the slightly wider
spectral distribution of the stars that peaks at about $1\,\ev$ with a
tipple power-law. We find the best fit spectra for the SD in a
cluster through:
\begin{eqnarray}
  \eph^2\frac{\dd^2 N_\ph}{\dd \eph \dd V}
  &\equiv& u_\sd(\eph) =  \sum_i N_i(\lx) u_i^\gal(\eph)\,,\nonumber \\ 
\rmn{where}\quad i&=&\{\rmn{stars,dust}\} \qquad \rmn{and}\\ 
% &=&  N_\dust(\lx) u_\dust^\gal(\eph) + N_\stars(\lx) u_\stars^\gal(\eph)  \nonumber \\ 
%  \\
  u_\stars^\gal(\eph) &=& \frac{23\,\rmn{eV}}{\rmn{cm}^3} 
  \left(\frac{1.23\,\rmn{eV}}{\eph}\right)^{1.9} \nonumber \\
  &\times&\left[1+\left(\frac{2.04\,\rmn{eV}}{\eph}\right)^{20}\right]
  ^{-\frac{1.9}{20}}\nonumber \\
  &\times& \left[1+\left(\frac{0.78\,\rmn{eV}}{\eph}\right)^{20}\right]^{-\frac{2.6}{20}}\,, \\
  u_\dust^\gal(\eph) &=& 
  \frac{40\,\rmn{eV}}{\rmn{cm}^3} 
  \left(\frac{0.0144\,\rmn{eV}}{\eph}\right)^{4.9}\nonumber \\
  &\times& \left[1+\left(\frac{0.0144\,\rmn{eV}}{\eph}\right)^{1.9}\right]^{-4.9}\,.
\end{eqnarray}

The normalization constant for the starlight $N_\stars(\lx)$ is derived
in the following three steps: (1) We use the average apparent magnitude
in the $(r+i)$-band $m_{(r+i),0.25}\approx 15.5$
\cite{2005MNRAS.358..949Z} of a cluster with the average mass
$\mvir=4.0\times10^{13}\msun$ at a redshift $z=0.25$ to derive the
luminosity from the stars:
\begin{equation}
L_\stars=10^{\frac{m_\odot-m_{(r+i),0.25}}{2.5}}
\left(\frac{D_\clu}{D_\odot}\right)^2 L_\odot
\approx 7.1\times10^{44}\,\rmn{erg}\,\rmn{s}^{-1}\,,
\label{eq:L_star}
\end{equation}
where the apparent magnitude of the sun $m_\odot\approx -27.1$, the
distance to the sun $D_\odot=4.85\times10^{-6}\,\rmn{pc}$, the
luminosity of the sun $L_\odot=3.85\times10^{33}\,\rmn{erg\,s}^{-1}$,
and the distance to the cluster $D_\clu=1.26\times10^9\,\rmn{pc}$. (2)
We use the IR to X-ray scaling relation in Eq.~(\ref{eq:SD-X}) to derive
a more general form of the luminosity from stars in
Eq.~(\ref{eq:L_star}) that is valid for an arbitrary cluster mass. The
cluster X-ray luminosities are found in the extended HIFLUGCS
catalogue \cite{2002ApJ...567..716R} to which we have added the Virgo
cluster. (3) The normalization constant for the stars is
determined from the above steps together with Eq.~(\ref{eq:E_SD})
where we integrate over the cluster volume and spectral distribution
of the stars. It is given by
\begin{equation}
 N_\stars(\lx) = 1.2\times10^{-9}\,
\left(\frac{\lx}{10^{44}\rmn{erg\,s}^{-1}}\right)^{0.53}\,.
\label{eq:N_stars}
\end{equation}
Note that since the luminosities include the contribution from
galaxies, we have integrated the spatial distribution including
galaxies to derive the normalization. Once the normalization is fixed,
the SD energy densities are derived from the spatial distribution
where galaxies have been excluded.

%***********************************************
